---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ckutils

<!-- badges: start -->
<!-- badges: end -->

The goal of ckutils is to provide some utilities for close-kin study design and implementation.

## Installation

You can install the development version of ckutils from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rasutt/ckutils")
```

## Example

This is a basic example which shows you how to simulate a wildlife population and a sampling-survey study of it.

### Simulation years and expected population size

Close-kin studies use the probabilities of close family relationships between pairs of sampled animals (close-kin pairs). These relationships are defined by the breeding events that occur prior to sampling, so ckutils lets you specify the number of years to simulate the population prior to the final survey year. This needs to be long enough that most of the relevant breeding events occur during the simulation.

ckutils simulates populations in which the expected population size grows by a constant factor called the population growth rate $\lambda$. Specifying this rate along with the expected population size $N$ in a reference year defines the expected population size in all years. It is good to check these as very small population sizes can be unsustainable, and very large ones slow to simulate.

```{r example}
# Load close-kin utilities package
library(ckutils)

# Survey years and simulation length
srvy_yrs = c(2010, 2015, 2020)
sim_len = 50

# Population growth rate, and population size in reference year
lambda = 1.05
ref_yr = 2010
exp_N_ref = 1500

# Simulation years
fnl_yr = tail(srvy_yrs, 1)
init_yr = fnl_yr - sim_len + 1
sim_yrs = init_yr:fnl_yr

# Expected population size over simulation years
exp_N_init = exp_N_ref * lambda^(init_yr - ref_yr)
exp_N_t = exp_N_init * lambda^(0:(sim_len - 1))

# Plot expected population size over simulation years
plot_exp_pop(sim_yrs, exp_N_t, ref_yr, exp_N_ref, srvy_yrs)
```

### Predicted numbers of close-kin pairs

ckutils simulates populations in which each animal survives from one year to the next with the same constant probability $\phi$, and becomes sexually mature at the same age $\alpha$. Each animal also has the same constant probability $p$ of being sampled in any survey (given that it is alive in that year).

These parameters, along with the population and survey parameters above, allow for approximations of the expected numbers of various close-kin pairs among sampled animals. A sufficient number of such pairs is essential to a successful study.

```{r}
# Annual survival rate and age of sexual maturity
phi = 0.9
alpha = 8

# Capture probability
p = 0.1

# Survey year indices
s_yr_inds = seq_along(srvy_yrs)

# Number of surveys
k = length(srvy_yrs)

# Per capita birth rate
rho = lambda - phi

# Find expected numbers of kin-pairs in population
find_exp_ns_kps(exp_N_t, s_yr_inds, phi, rho, lambda, alpha, srvy_yrs, k)
```

### Simulate population and study



```{r}
# Number of loci in genome
L = 100

# Initial minor allele frequency
imaf = 0.5

# Set random seed for testing
set.seed(1)

# Simulate one population and study
pop_study = sim_pop_study(
  phi, lambda, exp_N_init, sim_len, srvy_yrs, k,
  fnl_yr, p, L, imaf, clvng.p = 0, tmp.emgn = 0,
  alpha = alpha, clvng.ints = F
)

# Look at it
head(pop_study)
```
```{r}
names(attributes(pop_study))
```
